---
title: "LPNF Boundary and Trails/Roads"
---

This notebook selects the LPNF boundary from a geodatabase of all California national forests and saves the entire boundary, the entire boundary buffered, just the northern section of the boundary, just the southern section of the boundary, and the buffered northern and southern sections.

Load packages
```{r}
library(tidyverse)
library(sf)
library(here)

# Define base directory file path
base_dir <- here::here("~/Library/CloudStorage/Box-Box/MEDS-SBBG-milkweed")
```

## Read in CA National Forest boundaries
```{r}
# California National Forest boundaries
boundary <- st_read(here::here(base_dir, "raw_data", "S_USA_AdministrativeForest.gdb"))
```

## Filter boundaries to Los Padres National Forest
```{r}
lpnf_boundary <- boundary %>% 
  filter(FORESTNAME %in% c("Los Padres National Forest")) %>%
  st_transform(crs = st_crs("EPSG:4326"))

# save a buffered boundary for species distribution modeling
forest_buffer <- lpnf_boundary  %>%
  st_buffer(dist = 1000)

# save boundaries as shapefiles

# st_write(lpnf_boundary,
#          here(base_dir, "clean_data", "lpnf_boundary", "lpnf_boundary", "lpnf_boundary.shp"),
#          append = FALSE)
# st_write(forest_buffer,
#         here(base_dir, "clean_data", "lpnf_boundary", "lpnf_boundary_buffered", "lpnf_boundary_buffered.shp"),
#         append = FALSE)
 
write_rds(lpnf_boundary, here(base_dir, "outputs","dashboard", "lpnf_boundary.rda"))
```

## Separate LPNF boundary into northern and southern sections

Finding max-y for southern section by visualizing
```{r}
ggplot() +
  geom_sf(data = forest_buffer)

# select 35.6 as max-y for subsection of southern portion! 
```

```{r}
# Create a bounding box for the south and north
bbox_south <- st_bbox(c(xmin = -118, ymin = 34.2, xmax = -121, ymax = 35.5),  crs = st_crs(4326))
bbox_north <- st_bbox(c(xmin = -122, ymin = 35.5, xmax = -121, ymax = 36.5),  crs = st_crs(4326))

# Crop boundary to the bounding boxes
lpnf_south <- st_crop(lpnf_boundary, bbox_south) %>%
  select(SHAPE)
lpnf_north <- st_crop(lpnf_boundary, bbox_north) %>% 
  select(SHAPE)

# create buffered boundaries for south and north
lpnf_south_buffered <- lpnf_south %>%
  st_buffer(dist = 1000)
lpnf_north_buffered <- lpnf_north %>%
  st_buffer(dist = 1000)
```

Plot to verify 
```{r}
lpnf_south_df <- lpnf_south %>%
  st_coordinates() %>%
  data.frame()

ggplot(lpnf_south_df) +
  geom_polygon(aes(x = X, y = Y))

ggplot() +
  geom_sf(data = lpnf_south, color = "green4")
```

Save boundaries as shapefiles
```{r}
# write the forest boundaries
# st_write(lpnf_north, here(base_dir, "clean_data", "lpnf_boundary", "lpnf_boundary_north", "lpnf_boundary_north.shp"), append = FALSE)
# st_write(lpnf_south, here(base_dir, "clean_data", "lpnf_boundary", "lpnf_boundary_south", "lpnf_boundary_south.shp"), append = FALSE)
# st_write(lpnf_north_buffered, here(base_dir, "clean_data", "lpnf_boundary", "lpnf_boundary_north_buffered", "lpnf_boundary_north_buffered.shp"), append = FALSE)
# st_write(lpnf_south_buffered, here(base_dir, "clean_data", "lpnf_boundary", "lpnf_boundary_south_buffered", "lpnf_boundary_south_buffered.shp"), append = FALSE)
```

