---
title: "Solar Radiation"
format: html
---

```{r}
library(terra)
```

```{r}
dem <- rast(here("~/../../capstone/milkweedmod/data/dem/lpnf_dem.tif"))
slope <- raster(here::here("~/../../capstone/milkweedmod/data/dem/lpnf_slope.tif"))
aspect <- raster(here::here("~/../../capstone/milkweedmod/data/dem/lpnf_aspect.tif"))
```

“Northness”: Cosine of aspect x sine of slope
- If facing due south, cosine of aspect = 1
- As slope gets steeper, sine of slope increases
- South facing vertical wall has a northness of -1
- If want to change to “southness” multiply by -1 for positive values
- North facing vertical wall has a northness of 1

```{r}
northness <- cos(aspect) * sin(slope)
```

“eastness” (sine(aspect) x sine(slope))
- If due east on a vertical wall –> 1

```{r}
eastness <- sin(aspect) * sin(slope)
```

Notes from cos/sin documentation:
- Angles are in radians, not degrees, for the standard versions (i.e., a right angle is π/2), and in ‘half-rotations’ for cospi etc.
- cospi(x), sinpi(x), and tanpi(x) are accurate for x values which are multiples of a half.

