---
title: "Identify Priority Sites"
date: "Last Updated: 2024-06-03"
---

This notebook joins the habitat suitability maps and the survey site accessibility index to create a survey site priority index for each of the four early-season milkweed species.

```{r setup}
# Load packages
library(terra)
library(here)
library(wallace)
library(leaflet)

# Define base directory file path
base_dir <- here::here("~/Library/CloudStorage/Box-Box/MEDS-SBBG-milkweed")
outputs_dir <- here::here(base_dir, "outputs")
```

# Read in Habitat Suitability Maps, Accessibility Index, Template Raster

```{r}
# ---- Template Raster ----
template_rast <- rast(here(base_dir, "clean_data", "site_accessibility", "template_raster.tif"))

# ---- Accessibility Index ----
access_index <- rast(here(outputs_dir, "site_accessibility_outputs", "access_index_final.tif"))


# ---- Habitat Suitability Maps ----

# Californica
californica <- rast(here(outputs_dir, "sdm_outputs", "californica_sdm.tif"))

# Eriocarpa
eriocarpa <- rast(here(outputs_dir, "sdm_outputs", "eriocarpa_sdm.tif"))

# Erosa
erosa <- rast(here(outputs_dir, "sdm_outputs", "erosa_sdm.tif"))

# Vestita
vestita <- rast(here(outputs_dir, "sdm_outputs" "vestita_sdm.tif"))

```


## Check compatibility 

```{r}
print("Do the habitat suitability maps have the same origin as the template raster?")
origin(californica) == origin(template_rast)


print("Does the accessibility index have the same origin as the template raster?")
origin(access_index) == origin(template_rast)

```


# Reproject Suitability Rasters

```{r}
californica_reproject <- project(californica, template_rast) # default method is bilinear interpolation, which is fine for this case
eriocarpa_reproject <- project(eriocarpa, template_rast)
erosa_reproject <- project(erosa, template_rast)
vestita_reproject <- project(vestita, template_rast)

print("Do the habitat suitability maps have the same origin as the template raster?")
origin(californica_reproject) == origin(template_rast)


```


# Multiply Suitability Maps and Accessibility Index to find Priority Sites


```{r}
# make priority sites index 

californica_priority <- californica_reproject * access_index
eriocarpa_priority <- eriocarpa_reproject * access_index
erosa_priority <- erosa_reproject * access_index
vestita_priority <- vestita_reproject * access_index


plot(californica_priority)
plot(eriocarpa_priority)

```


# Rescale to make Priority Index from 0-1

Lowest priority: 0
Highest priority: 1

```{r}
# Source rescale raster function from scripts folder
source(here("R", "rescale_raster.R"))

# Rescale priority for each species of milkweed

californica_rescaled <- rescale_raster(californica_priority)
eriocarpa_rescaled <- rescale_raster(eriocarpa_priority)
erosa_rescaled <- rescale_raster(erosa_priority)
vestita_rescaled <- rescale_raster(vestita_priority)

# Plot to check output
plot(californica_rescaled)
plot(eriocarpa_rescaled)
plot(erosa_rescaled)
plot(vestita_rescaled)
```

## rename layer names
```{r}
names(californica_rescaled) <- "californica_priority"
names(eriocarpa_rescaled) <- "eriocarpa_priority"
names(erosa_rescaled) <- "erosa_priority"
names(vestita_rescaled) <- "vestita_priority" 
```


# Save outputs to outputs folder

```{r}
#writeRaster(californica_rescaled, here(outputs_dir, "priority_sites_outputs", "californica_priority.tif"), overwrite = TRUE)
# 
#writeRaster(eriocarpa_rescaled, here(outputs_dir, "priority_sites_outputs", "eriocarpa_priority.tif"), overwrite = TRUE)
# 
#writeRaster(erosa_rescaled, here(outputs_dir, "priority_sites_outputs", "erosa_priority.tif"), overwrite = TRUE)
# 
#writeRaster(vestita_rescaled, here(outputs_dir, "priority_sites_outputs", "vestita_priority.tif"), overwrite = TRUE)
```

# Save priority maps to GitHub

```{r}
# writeRaster(californica_rescaled, here("outputs", "dashboard", "californica_priority.tif"), overwrite = TRUE)
# 
# writeRaster(eriocarpa_rescaled, here("outputs", "dashboard", "eriocarpa_priority.tif"), overwrite = TRUE)
# 
# writeRaster(erosa_rescaled, here("outputs", "dashboard", "erosa_priority.tif"), overwrite = TRUE)
# 
# writeRaster(vestita_rescaled, here("outputs", "dashboard", "vestita_priority.tif"), overwrite = TRUE)
```

# Plot with leaflet

```{r}
source(here("R", "addLegend_decreasing.R"))

californica <- rast(here("outputs", "dashboard", "californica_priority.tif"))
eriocarpa <- rast(here("outputs", "dashboard", "eriocarpa_priority.tif"))
erosa <- rast(here("outputs", "dashboard", "erosa_priority.tif"))
vestita <- rast(here("outputs", "dashboard", "vestita_priority.tif"))
lpnf_boundary <- st_read(here("~/../../capstone/milkweedmod/clean_data/lpnf_boundary/lpnf_boundary/lpnf_boundary.shp"),
                         quiet = TRUE)

# define color palette 
pal_priority <- c("#FFFFFF", "#E1CCEF", "#C399DF", "#A666D0", "#8833C0", "#6A00B0")

# ---------------------californica---------------------
# get raster values
californicaVals <- getRasterVals(californica) # change for different types

legendPal <- colorNumeric(pal_priority, californicaVals, na.color = 'transparent')
rasPal <- colorNumeric(pal_priority, californicaVals, na.color = 'transparent')

californica_priority_leaflet <- leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
  
  addProviderTiles(providers$Esri.WorldTerrain) %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = californicaVals,
                       labFormat = reverseLabel(), decreasing = TRUE,
                       title = "Priority Index:<br><em>Asclepias californica</em><br>") %>%
  # map model prediction raster and transfer polygon
  addRasterImage(californica,
                 colors = rasPal,
                 method = "ngb") %>%
  
  # #add transfer polygon (user drawn area)
  addPolygons(data = lpnf_boundary, fill = FALSE,
              weight = 2, color = "black", group = 'xfer') %>% 
  addScaleBar(position = "bottomright")

californica_priority_leaflet


# ---------------------eriocarpa---------------------

# get raster values
eriocarpaVals <- getRasterVals(eriocarpa) # change for different types

legendPal <- colorNumeric(pal_priority, eriocarpaVals, na.color = 'transparent')
rasPal <- colorNumeric(pal_priority, eriocarpaVals, na.color = 'transparent')

eriocarpa_priority_leaflet <- leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
  
  addProviderTiles(providers$Esri.WorldTerrain) %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = eriocarpaVals,
                       labFormat = reverseLabel(), decreasing = TRUE,
                       title = "Priority Index:<br><em>Asclepias eriocarpa</em><br>") %>%
  # map model prediction raster and transfer polygon
  addRasterImage(eriocarpa,
                 colors = rasPal,
                 method = "ngb") %>%
  
  # #add transfer polygon (user drawn area)
  addPolygons(data = lpnf_boundary, fill = FALSE,
              weight = 2, color = "black", group = 'xfer') %>% 
  addScaleBar(position = "bottomright")

eriocarpa_priority_leaflet


# ---------------------erosa---------------------

# get raster values
erosaVals <- getRasterVals(erosa) # change for different types

legendPal <- colorNumeric(pal_priority, erosaVals, na.color = 'transparent')
rasPal <- colorNumeric(pal_priority, erosaVals, na.color = 'transparent')

erosa_priority_leaflet <- leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
  
  addProviderTiles(providers$Esri.WorldTerrain) %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = erosaVals,
                       labFormat = reverseLabel(), decreasing = TRUE,
                       title = "Priority Index:<br><em>Asclepias erosa</em><br>") %>%
  # map model prediction raster and transfer polygon
  addRasterImage(erosa,
                 colors = rasPal,
                 method = "ngb") %>%
  
  # #add transfer polygon (user drawn area)
  addPolygons(data = lpnf_boundary, fill = FALSE,
              weight = 2, color = "black", group = 'xfer') %>% 
  addScaleBar(position = "bottomright")

erosa_priority_leaflet

# ---------------------vestita---------------------

# get raster values
vestitaVals <- getRasterVals(vestita) # change for different types

legendPal <- colorNumeric(pal_priority, vestitaVals, na.color = 'transparent')
rasPal <- colorNumeric(pal_priority, vestitaVals, na.color = 'transparent')

vestita_priority_leaflet <- leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
  
  addProviderTiles(providers$Esri.WorldTerrain) %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = vestitaVals,
                       labFormat = reverseLabel(), decreasing = TRUE,
                       title = "Priority Index:<br><em>Asclepias vestita</em><br>") %>%
  # map model prediction raster and transfer polygon
  addRasterImage(vestita,
                 colors = rasPal,
                 method = "ngb") %>%
  
  # #add transfer polygon (user drawn area)
  addPolygons(data = lpnf_boundary, fill = FALSE,
              weight = 2, color = "black", group = 'xfer') %>% 
  addScaleBar(position = "bottomright")

vestita_priority_leaflet
```

