---
title: "Making a template raster for site accessibility data preparation"
date_created: "2024-17-04"
author: "Amanda Herbst"
---
```{r}
library(tidyverse)
library(terra)
library(here)
```

We have been using a template raster based on the SDM output, but wanted to create a template from scratch so we are not dependent on the model output.

Currently basing this template on `outputs/sdm_outputs/californica_bioclim_dem.tif` in Taylor (4/17/24).

- CRS: EPSG:4326
- Resolution: 0.008333333 x 0.008333333
- Extent: -122.01324133097, -117.479907997637, 33.679743833477, 36.9214105001436 (xmin, xmax, ymin, ymax)
- Origin: 0.003425336, -0.003589500
- number of rows: 389
- Number of columns: 544

## Build template
```{r}
# # define resolution and corners of the extent
# res <- c(0.008333333, 0.008333333)
# xmin <- -122.01324133097
# xmax <- -117.479907997637
# ymin <- 33.679743833477
# ymax <- 36.9214105001436
# crs <- "EPSG:4326"
#   
# temp_rast <- rast(nrow = 389, ncol = 544, resolution = res,
#                   xmin = xmin, xmax = xmax,
#                   ymin = ymin, ymax = ymax,
#                   vals = 0, crs = crs)
# # plot to verify
# plot(temp_rast)
```

## Write to csv
```{r}
# writeRaster(temp_rast, 
#             here("~/../../capstone/milkweedmod/clean_data/site_accessibility/template_raster.tif"))
```

# 4.21.24 

The origin of the above template is slightly off from the origin of the model raster, so instead of going off the exact values of the model raster, I am going to define similar values. The problem seems to come from the resolution that R prints out with 6 repeating 3s, but I believe are infinite repeating 3s. This adjusted template will be used for all accessibility calculations and then then model output will be adjusted to the same resolution/extent before combining for final high-priority sites.

New values: 
- CRS: EPSG:4326 (same)
- Resolution: 0.0083 x 0.0083 (new)
- Extent: -122.0, -117.5, 33.6, 36.9 (xmin, xmax, ymin, ymax)
- Origin: 0.0017, 0.0016
- number of rows: 389
- Number of columns: 544


```{r}
# define resolution and corners of the extent
res <- c(0.0083, 0.0083)
xmin <- -122.0
xmax <- -117.5
ymin <- 33.6
ymax <- 36.9
crs <- "EPSG:4326"
  
temp_rast <- rast(nrow = 389, ncol = 544, resolution = res,
                  xmin = xmin, xmax = xmax,
                  ymin = ymin, ymax = ymax,
                  vals = 0, crs = crs)
# plot to verify
plot(temp_rast)

# check origin
origin(temp_rast)
```

```{r}
# writeRaster(temp_rast,
#             here("~/../../capstone/milkweedmod/clean_data/site_accessibility/template_raster.tif"),
#             overwrite = TRUE)
```
