---
title: "Making a template raster for site accessibility data preparation"
date_created: "2024-06-04"
---

This notebook creates a template raster which is used to reproject and rasterize individual layers as needed in preparation for creating the survey site accessibility index.

```{r}
library(tidyverse)
library(terra)
library(here)

# Define base directory file path
base_dir <- here("~/Library/CloudStorage/Box-Box/MEDS-SBBG-milkweed")

template_dir <- here(base_dir, "clean_data", "site_accessibility")
```

We have been using a template raster based on the SDM output, but wanted to create a template from scratch so we are not dependent on the model output. This is especially helpful to make because the resolution of the model outputs is 0.00833 repeating, which is not ideal and impossible to make rasters with that same exact resolution. This adjusted template will be used for all accessibility calculations and then then model output will be adjusted to the same resolution/extent before combining for final high-priority sites.

Basing this template on model outputs:

- CRS: EPSG:4326
- Resolution: 0.008333333 x 0.008333333
- Extent: -122.01324133097, -117.479907997637, 33.679743833477, 36.9214105001436 (xmin, xmax, ymin, ymax)
- Origin: 0.003425336, -0.003589500
- number of rows: 389
- Number of columns: 544

# Build template 

New values: 
- CRS: EPSG:4326 (same)
- Resolution: 0.0083 x 0.0083 (new)
- Extent: -122.0, -117.5, 33.6, 36.9 (xmin, xmax, ymin, ymax)
- Origin: 0.0017, 0.0016
- number of rows: 389
- Number of columns: 544


```{r}
# define resolution and corners of the extent
res <- c(0.0083, 0.0083)
xmin <- -122.0
xmax <- -117.5
ymin <- 33.6
ymax <- 36.9
crs <- "EPSG:4326"
  
temp_rast <- rast(nrow = 389, ncol = 544, resolution = res,
                  xmin = xmin, xmax = xmax,
                  ymin = ymin, ymax = ymax,
                  vals = 0, crs = crs)
# plot to verify
plot(temp_rast)

# check origin
origin(temp_rast)
```


# Write to box folder

```{r}
# writeRaster(temp_rast,
#             here(template_dir, "template_raster.tif"),
#             overwrite = TRUE)
```
