# Intro

```{r}
library(sf)
library(tidyverse)
```

# SDM 
## Add new data

### Read in and clean new milkweed survey data
```{r}
# read in properly formatted milkweed polygon data
new_milkweed_raw <- st_read("path/to/data")
```
Now that the new data is read in, we want to check that the structure of the data looks correct. Using `View(new_milkweed_raw)` look for the columns in the data `Milkweed_P`, `MilkweedSp`, and `geometry` (this will be the very last column). These are the columns we want to use for the species distribution modeling. 
```{r}
# view the data frame; the data frame will pop up in a new tab at the top
View(new_milkweed_raw)
```

##### Troubleshooting Errors
Problems with reading the the data:

Different column names/missing columns:

### Clean New Data
If everything is looking good, let's move on to cleaning the data up a bit. We want to select only the columns we need for modeling, and rename them to names that are a bit simpler. Since this data has both "yes" and "no" values in the `Milkweed_P`, we want to filter to only keep the "yes" values. Finally, let's transform the coordinate reference system (CRS) to EPSG:4326, since this the the CRS that the prepared environmental data is in. 
```{r}
milkweed_clean <- new_milkweed_raw |> 
  janitor::clean_names() |> 
  filter(milkweed_p != "no")  %>%
  st_transform("EPSG:4326") %>%
  dplyr::select(milkweed_sp)

# check that the new clean data is looking correct
head(milkweed_clean)
```

Here's what the data should look like:
```
       milkweed_sp                       geometry
1 Asclepias vestita MULTIPOLYGON (((-119.3003 3...
2 Asclepias erosa   MULTIPOLYGON (((-119.1852 3...
3 Asclepias cali... MULTIPOLYGON (((-119.402 34...
```

##### Troubleshooting Errors


## Model






## Updating “closed” Trails and Roads
Direct to trails and roads doc that already has commented note on how to filter (note again that this is only applicable to the south)

### Using those updates to update site accessibility 
Instruct on where/how to run rescale layers, etc. again.

## Updating Site Priority 
Using new outputs from SDM – north and south, species-specific and max(all) 

Save outputs to _____ (note here that this is where the dashboard will be pulling from) 


