---
title: "Site Accessibility Index Score"
author: "Amanda Herbst"
date-created: "2024-02-15"
---

Our goal is to create a site accessibility index score that indicates areas in the LPNF that SBBG researchers can access easily. We will base this index on distance from trails/roads, public/private land, vegetation density, and slope.

There are tools for calculating this in ArcGIS but not much has been found in R.

**Land Ownership**: [ca.gov](https://gis.data.ca.gov/datasets/f73858e200634ca888b19ca8c78e3aed_0/explore?location=34.536765%2C-119.257630%2C7.00)

# Load packages

```{r}
library(tidyverse)
library(sf)
```

# Load lpnf data for bounding box

```{r}
boundary <- st_read(here("~/../../capstone/milkweedmod/data/lpnf_boundary_data/S_USA_AdministrativeForest.gdb/"))
lpnf_boundary <- boundary %>% 
  filter(FORESTNAME %in% c("Los Padres National Forest"))
lpnf_bb <- st_bbox(lpnf_boundary)

st_crs(lpnf_boundary)
```

# Load data

```{r}
# land ownership ----

land_ownership <- st_read(here("~/../../capstone/milkweedmod/data/California_Land_Ownership/ownership23_1.shp"))
st_crs(land_ownership)

# reproject 
ownership_reprojected <- land_ownership %>%  st_transform(crs = st_crs(lpnf_boundary))
st_crs(ownership_reprojected)
```

# Explore data

```{r}
ggplot() +
  geom_sf(data = ownership_reprojected, aes(fill = OWN_LEVEL)) +
  geom_sf(data = lpnf_boundary,
          color = "darkgreen",
          fill = "transparent",
          lwd = 1) +
  coord_sf(xlim = c(lpnf_bb[1], lpnf_bb[3]),
           ylim = c(lpnf_bb[2], lpnf_bb[4])) +
  theme_bw()
```
