---
title: "erosa: Bioclim, canopy, northness, eastness"
format: html
---

This notebook contains code for running a MaxEnt SDM for *A. erosa* within the Los Padres National Forest boundary, model selection, model transfer, and mapping. 

```{r}
library(spocc)
library(spThin)
library(dismo)
library(sf)
library(ENMeval)
library(wallace)
library(tidyverse)
library(here)
library(rgbif)
source(here("R", "addLegend_decreasing.R"))
source(here("R", "setup.R"))
```

## Analysis

### Obtain occurrence data for A. erosa

```{r}
# rename milkweed occurrence data
 occs_Ac <- erosa_points
```

```{r}
# download GBIF occurrence data for this species; this takes time if there are many data points!
gbif_data <- occ_data(scientificName = "Asclepias erosa", hasCoordinate = TRUE, limit = 300000) 
occs_Ac <- gbif_data$data %>%
  filter(stateProvince == "California",
         decimalLongitude < -118.8, 
         decimalLatitude < 34.9) %>%
  select(decimalLongitude, decimalLatitude, scientificName) %>%
  rename(longitude = decimalLongitude,
         latitude = decimalLatitude,
         scientific_name = scientificName) %>%
  mutate(occID = row_number(),
         scientific_name = "Asclepias erosa") %>%
  rbind(erosa_points)

# ggplot() + 
#   geom_point(data = occs_Ac, aes(x = longitude, y = latitude), size = 0.5) + 
#   geom_sf(data = lpnf_south_buffered, alpha = 0.5)

# # map the occurrence data:
# library(maptools)
# data(wrld_simpl)
# plot(wrld_simpl, axes=TRUE, col="light yellow")
# # add the points
# points(gbif_erosa$decimalLongitude, gbif_erosa$decimalLatitude, col='red', pch=20, cex=0.75)
```


### Obtain environmental data

```{r}
occs_geom_Ac <- occs_Ac[c("longitude", "latitude")]
occs_vals_Ac <- as.data.frame(raster::extract(envs_Ac, occs_geom_Ac, cellnumbers = TRUE))

# add columns for env variable values for each occurrence record
occs_Ac <- cbind(occs_Ac, occs_vals_Ac)
```


### Process Occurrence Data

```{r}
# Thin occurrences to 0.05 km
occs_Ac <- poccs_thinOccs(
  occs = occs_Ac, 
  thinDist = 0.01) # adjust this value if needed
```

### Process environmental data

Sampling background points and corresponding environmental data for the southern lpnf boundary.

```{r}
# Generate background extent
bgExt_Ac <- lpnf_south_buffered

# Mask environmental data to provided extent
bgMask_Ac <- penvs_bgMask(
  occs = occs_Ac,
  envs = envs_Ac,
  bgExt = bgExt_Ac)
# Sample background points from the provided area
bgSample_Ac <- penvs_bgSample(
  occs = occs_Ac,
  bgMask =  bgMask_Ac,
  bgPtsNum = 5000)
# Extract values of environmental layers for each background point
bgEnvsVals_Ac <- as.data.frame(raster::extract(bgMask_Ac,  bgSample_Ac))
##Add extracted values to background points table
bgEnvsVals_Ac <- cbind(scientific_name = paste0("bg_", "Asclepias erosa"), bgSample_Ac,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Ac)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using block, a spatial partition method.

```{r}
# Partitioned data
groups_Ac <- part_partitionOccs(
  occs = occs_Ac ,
  bg =  bgSample_Ac, 
  method = "block") 
```


### Build and Evaluate Niche Model

Generating a species distribution model using the maxnet algorithm as
implemented in ENMeval V2.0 (with clamping = FALSE). For tuning using L,
LQ, H, LQH, LQHP feature classes and regularization multipliers in the
1, 2 range increasing by 1. Not using any categorical predictor
variables.

```{r}
# Run maxent model for the selected species
model_Ac <- model_maxent(
  occs = occs_Ac,
  bg = bgEnvsVals_Ac,
  user.grp = groups_Ac, 
  bgMsk = bgMask_Ac,
  rms = c(0.5, 4), 
  rmsStep =  0.5,
  fcs = c('L', 'LQ'),
  clampSel = FALSE,
  algMaxent = "maxnet",
  parallel = FALSE,
  numCores = 7)
```

https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html#different-parameterizations
```{r}
model_Ac@results %>%
  gt::gt()
```

### Visualize

Generate a map of the maxnet generated model with no threshold

```{r}
# Select current model and obtain raster prediction
m_Ac <- model_Ac@models[["fc.LQ_rm.0.5"]] # change this for different models
predSel_Ac <- predictMaxnet(m_Ac, bgMask_Ac,
                                          type = "cloglog", # change for different types
                                          clamp = FALSE)
#Get values of prediction
mapPredVals_Ac <- getRasterVals(predSel_Ac, "cloglog") # change for different types

#Define colors and legend  
rasCols <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")

legendPal <- colorNumeric(rasCols, mapPredVals_Ac, na.color = 'transparent')
rasPal <- colorNumeric(rasCols, mapPredVals_Ac, na.color = 'transparent')

#Generate map
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) 
m  %>%
  addCircleMarkers(data = occs_Ac, lat = ~latitude, lng = ~longitude,
                   radius = 2, color = 'black', fill = TRUE, fillColor = "black",
                   fillOpacity = 0.2, weight = 2) %>% 
  ##Add model prediction
  addRasterImage(predSel_Ac, colors = rasPal, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
  leaflet::addLegend("bottomright", pal = legendPal, values = mapPredVals_Ac, layerId = "train",
            labFormat = reverseLabel(),
            title = "Asclepias californica.<br>Predicted Suitability") %>%
  addPolygons(data = lpnf_south_buffered,
              fill = FALSE,
              color = "black",
              weight = 2)
```

## Model Transfer

```{r}
# Generate a transfer of the model to the desired area
xfer_area_Ac <- xfer_area(
  evalOut = model_Ac,
  curModel = "fc.LQ_rm.0.5",
  envs = envs_Ac, 
  outputType = "cloglog",
  alg = "maxnet",
  clamp = FALSE,
  xfExt = lpnf_north) 
# store the cropped transfer variables
xferExt_Ac <- xfer_area_Ac$envs_north2

#map result
mapXferVals_Ac <- getRasterVals(xfer_area_Ac$xferArea,"cloglog")
rasCols_Ac <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
# if no threshold specified
legendPal <- colorNumeric(rev(rasCols_Ac), mapXferVals_Ac, na.color = 'transparent')
rasPal_Ac <- colorNumeric(rasCols_Ac, mapXferVals_Ac, na.color = 'transparent')
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap)
m %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = mapXferVals_Ac, layerId = "xfer",
            labFormat = reverseLabel(), decreasing = FALSE,
            title = "Asclepias erosa<br>Predicted Suitability<br>(Transferred)") %>%
# map model prediction raster and transfer polygon
  clearMarkers() %>% clearShapes() %>% removeImage('xferRas') %>%
  addRasterImage(xfer_area_Ac$xferArea, colors = rasPal_Ac, opacity = 0.7,
                 layerId = 'xferRas', group = 'xfer', method = "ngb") %>%
 ##add transfer polygon (user drawn area)
  addPolygons(data = lpnf_north, fill = FALSE,
              weight = 2, color = "black", group = 'xfer')
```


# Plot both together

```{r}
leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = mapXferVals_Ac, layerId = "xfer",
            labFormat = reverseLabel(), decreasing = FALSE,
            title = "<em>Asclepias erosa</em><br>Predicted Suitability<br>") %>%
# map model prediction raster and transfer polygon
  clearMarkers() %>% clearShapes() %>% removeImage('xferRas') %>%
  addRasterImage(xfer_area_Ac$xferArea, colors = rasPal_Ac, opacity = 0.7,
                 layerId = 'xferRas', group = 'xfer', method = "ngb") %>%
 ##add transfer polygon (user drawn area)
  addPolygons(data = lpnf_north, fill = FALSE,
              weight = 2, color = "black", group = 'xfer')  %>%
  addCircleMarkers(data = occs_Ac, lat = ~latitude, lng = ~longitude,
                   radius = 2, color = 'black', fill = TRUE, fillColor = "black",
                   fillOpacity = 0.2, weight = 2) %>% 
  ##Add model prediction
  addRasterImage(predSel_Ac, colors = rasPal, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
  addPolygons(data = lpnf_south,
              fill = FALSE,
              color = "black",
              weight = 2)
```

```{r}
erosa_model <- merge(xfer_area_Ac$xferArea, predSel_Ac) %>%
  rast()
names(erosa_model) <- "fc.LQ_rm.0.5"
writeRaster(erosa_model, here("~/../../capstone/milkweedmod/outputs/sdm_outputs/erosa_sdm.tif"), overwrite=TRUE)
write_rds(model_Ac, here("~/../../capstone/milkweedmod/outputs/sdm_outputs/erosa_sdm.rda"))
writeRaster(erosa_model, "outputs/dashboard/erosa_sdm.tif")
```

```{r}
erosa_pred <- rast(here("~/../../capstone/milkweedmod/outputs/sdm_outputs/erosa_sdm.tif")) %>%
  mask(lpnf_boundary)
pal_habitat <- c("#FFFFFF", "#EFCCCC", "#DF9999", "#D06666", "#C03333", "#B00000")
mapPredVals_erosa <- getRasterVals(erosa_pred) # change for different types

legendPal_erosa <- colorNumeric(pal_habitat, mapPredVals_erosa, na.color = 'transparent')
rasPal_erosa <- colorNumeric(pal_habitat, mapPredVals_erosa, na.color = 'transparent')

leaflet(options = leafletOptions(zoomControl = FALSE)) %>% 
  addProviderTiles(providers$Esri.WorldTerrain) %>%
  addLegend_decreasing("bottomleft", pal = legendPal_erosa, values = mapPredVals_erosa,
            labFormat = reverseLabel(), decreasing = TRUE,
            title = "<em>Asclepias erosa</em><br>Predicted Suitability<br>") %>%
 ##add transfer polygon (user drawn area)
  addPolygons(data = lpnf_north, fill = FALSE,
              weight = 2, color = "black")  %>%
  ##Add model prediction
  addRasterImage(erosa_pred, colors = rasPal_erosa, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
  addPolygons(data = lpnf_south,
              fill = FALSE,
              color = "black",
              weight = 2) %>%
  addScaleBar(position = "bottomright")
```

