---
title: "eriocarpa: Bioclim, canopy, aspect, slope"
format: html
---

```{r}
library(spocc)
library(spThin)
library(dismo)
library(sf)
library(ENMeval)
library(wallace)
library(tidyverse)
source("scripts/addLegend_decreasing.R")
```

```{r}
# Prepare milkweed data 
milkweed_data_raw <- st_read(here::here("~/../../capstone/milkweedmod/data/milkweed_polygon_data/"))

# Read in bioclim, canopy, slope, and aspect data
bioclim <- brick(here::here("~/../../capstone/milkweedmod/data/bioclim/wallace_bioclim.tif"))
canopy <- raster(here::here("~/../../capstone/milkweedmod/data/canopy_cover/canopy_cover_cleaned.tif"))
slope <- raster(here::here("~/../../capstone/milkweedmod/data/dem/lpnf_slope.tif"))
aspect <- raster(here::here("~/../../capstone/milkweedmod/data/dem/lpnf_aspect.tif"))

# California National Forest boundaries
lpnf_boundary <- st_read(here("~/../../capstone/milkweedmod/data/lpnf_boundary_data/lpnf_boundary/lpnf_boundary.shp"))
lpnf_north <- st_read(here("~/../../capstone/milkweedmod/data/lpnf_boundary_data/lpnf_boundary_north/lpnf_boundary_north.shp")) %>%
  st_buffer(dist = 1000)
lpnf_south <- st_read(here("~/../../capstone/milkweedmod/data/lpnf_boundary_data/lpnf_boundary_south/lpnf_boundary_south.shp")) %>%
  st_buffer(dist = 1000)


# remove "No" observations
milkweed_presence <- milkweed_data_raw |> 
  janitor::clean_names() |> 
  filter(milkweed_p != "no")  %>%
  st_transform("EPSG:4326") %>%
  filter(milkweed_sp == "Asclepias eriocarpa") %>%
  dplyr::select(milkweed_sp)

# make points
multi.p <- st_cast(milkweed_presence, "MULTIPOINT")
my_points <- multi.p  %>% st_cast("POINT")

milkweed_points <- my_points %>%
  st_coordinates() %>%
  data.frame() %>%
  mutate(scientific_name = "Asclepias californica") %>%
  mutate(occID = row_number()) %>%
  rename(longitude = X,
         latitude = Y)
```

##### Crop and stack bioclim canopy cover slope aspect

```{r}

# resoource: https://gis.stackexchange.com/questions/403536/raster-extent-issue-in-r
bioclim_resample <- resample(bioclim, canopy)
aspect_resample <- resample(aspect, canopy)
slope_resample <- resample(slope, canopy)

canopy_mask <- mask(canopy, bioclim_resample)
aspect_mask <- mask(aspect_resample, bioclim_resample)
slope_mask <- mask(slope_resample, bioclim_resample)

# stack
envs_Ac <- raster::stack(bioclim_resample, canopy_mask, aspect_mask, slope_mask)
crs(envs_Ac) <- "EPSG:4326"

envs_north <- crop(envs_Ac, lpnf_north)
```


## Analysis

### Obtain Occurrence Data

```{r}
# rename milkweed occurrence data
 occs_Ac <- milkweed_points # all species
```

### Obtain environmental data

Using WorldClim (<http://www.worldclim.org/>) bioclimatic dataset at
resolution of 0.5 arcmin.

```{r}
occs_geom_Ac <- occs_Ac[c("longitude", "latitude")]
occs_vals_Ac <- as.data.frame(raster::extract(envs_Ac, occs_geom_Ac, cellnumbers = TRUE))

# add columns for env variable values for each occurrence record
occs_Ac <- cbind(occs_Ac, occs_vals_Ac)
```


### Process Occurrence Data

Thinning the occurrences to 0.05 km

```{r}
# Thin occurrences 
occs_Ac <- poccs_thinOccs(
  occs = occs_Ac, 
  thinDist = 0.05) # adjust this value if needed
```

### Process environmental data

```{r}
# Generate background extent
bgExt_Ac <- lpnf_south

# Mask environmental data to provided extent
bgMask_Ac <- penvs_bgMask(
  occs = occs_Ac,
  envs = envs_Ac,
  bgExt = bgExt_Ac)
# Sample background points from the provided area
bgSample_Ac <- penvs_bgSample(
  occs = occs_Ac,
  bgMask =  bgMask_Ac,
  bgPtsNum = 10000)
# Extract values of environmental layers for each background point
bgEnvsVals_Ac <- as.data.frame(raster::extract(bgMask_Ac,  bgSample_Ac))
##Add extracted values to background points table
bgEnvsVals_Ac <- cbind(scientific_name = paste0("bg_", "Asclepias eriocarpa"), bgSample_Ac,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Ac)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using block, a spatial partition method.

```{r}
# R code to get partitioned data
groups_Ac <- part_partitionOccs(
  occs = occs_Ac ,
  bg =  bgSample_Ac, 
  method = "block") 
```


### Build and Evaluate Niche Model

Generating a species distribution model using the maxnet algorithm as
implemented in ENMeval V2.0 (with clamping = FALSE). For tuning using L,
LQ, H, LQH, LQHP feature classes and regularization multipliers in the
1, 2 range increasing by 1. Not using any categorical predictor
variables.

```{r}
# Run maxent model for the selected species
model_Ac <- model_maxent(
  occs = occs_Ac,
  bg = bgEnvsVals_Ac,
  user.grp = groups_Ac, 
  bgMsk = bgMask_Ac,
  rms = c(1, 2), 
  rmsStep =  1,
  fcs = c('L', 'LQ'),
  clampSel = FALSE,
  algMaxent = "maxnet",
  parallel = FALSE,
  numCores = 7)

# we can only run L and LQ; any models with H (hinge) are not what we want- get the weird coefficients error
```

https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html#different-parameterizations
```{r}
model_Ac@results %>%
  gt::gt()
```

### Visualize

Generate a map of the maxnet generated model with no threshold

```{r}
# Select current model and obtain raster prediction
m_Ac <- model_Ac@models[["fc.L_rm.2"]] # change this for different models
predSel_Ac <- predictMaxnet(m_Ac, bgMask_Ac,
                                          type = "cloglog", # change for different types
                                          clamp = FALSE)
#Get values of prediction
mapPredVals_Ac <- getRasterVals(predSel_Ac, "cloglog") # change for different types

#Define colors and legend  
rasCols <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")

legendPal <- colorNumeric(rasCols, mapPredVals_Ac, na.color = 'transparent')
rasPal <- colorNumeric(rasCols, mapPredVals_Ac, na.color = 'transparent')

#Generate map
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) 
m  %>%
  addCircleMarkers(data = occs_Ac, lat = ~latitude, lng = ~longitude,
                   radius = 2, color = 'black', fill = TRUE, fillColor = "black",
                   fillOpacity = 0.2, weight = 2) %>% 
  ##Add model prediction
  addRasterImage(predSel_Ac, colors = rasPal, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
  leaflet::addLegend("bottomright", pal = legendPal, values = mapPredVals_Ac, layerId = "train",
            labFormat = reverseLabel(),
            title = "Asclepias eriocarpa<br>Predicted Suitability") %>%
  addPolygons(data = lpnf_south,
              fill = FALSE,
              color = "black",
              weight = 2)
```

## Model Transfer

```{r}
# Generate a transfer of the model to the desired area
xfer_area_Ac <- xfer_area(
  evalOut = model_Ac,
  curModel = "fc.L_rm.2",
  envs = envs_north , 
  outputType = "cloglog",
  alg = "maxnet",
  clamp = FALSE,
  xfExt = lpnf_north) 
# store the cropped transfer variables
xferExt_Ac <- xfer_area_Ac$envs_north

#map result
mapXferVals_Ac <- getRasterVals(xfer_area_Ac$xferArea,"cloglog")
rasCols_Ac <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
# if no threshold specified
legendPal <- colorNumeric(rev(rasCols_Ac), mapXferVals_Ac, na.color = 'transparent')
rasPal_Ac <- colorNumeric(rasCols_Ac, mapXferVals_Ac, na.color = 'transparent')
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap)
m %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = mapXferVals_Ac, layerId = "xfer",
            labFormat = reverseLabel(), decreasing = FALSE,
            title = "Asclepias eriocarpa<br>Predicted Suitability<br>(Transferred)") %>%
# map model prediction raster and transfer polygon
  clearMarkers() %>% clearShapes() %>% removeImage('xferRas') %>%
  addRasterImage(xfer_area_Ac$xferArea, colors = rasPal_Ac, opacity = 0.7,
                 layerId = 'xferRas', group = 'xfer', method = "ngb") %>%
 ##add transfer polygon (user drawn area)
  addPolygons(data = lpnf_north, fill = FALSE,
              weight = 2, color = "black", group = 'xfer')
```


# Plot both together

```{r}
leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) %>%
  addLegend_decreasing("bottomleft", pal = legendPal, values = mapXferVals_Ac, layerId = "xfer",
            labFormat = reverseLabel(), decreasing = FALSE,
            title = "<em>Asclepias eriocarpa</em><br>Predicted Suitability<br>") %>%
# map model prediction raster and transfer polygon
  clearMarkers() %>% clearShapes() %>% removeImage('xferRas') %>%
  addRasterImage(xfer_area_Ac$xferArea, colors = rasPal_Ac, opacity = 0.7,
                 layerId = 'xferRas', group = 'xfer', method = "ngb") %>%
 ##add transfer polygon (user drawn area)
  addPolygons(data = lpnf_north, fill = FALSE,
              weight = 2, color = "black", group = 'xfer')  %>%
  addCircleMarkers(data = occs_Ac, lat = ~latitude, lng = ~longitude,
                   radius = 2, color = 'black', fill = TRUE, fillColor = "black",
                   fillOpacity = 0.2, weight = 2) %>% 
  ##Add model prediction
  addRasterImage(predSel_Ac, colors = rasPal, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
  addPolygons(data = lpnf_south,
              fill = FALSE,
              color = "black",
              weight = 2)
```

```{r}
dismo::response(model_Ac@models[[1]], data = envs_Ac)
```

